Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-08-18 survey/survey.tex
Fixed the use of * in idempotence.

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\sumtype}{+}
\newcommand{\prodtype}{\times}
\newcommand{\inleft}[1]{\textsf{left}~#1}
\newcommand{\inright}[1]{\textsf{right}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_{\sumtype}}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_{\sumtype}}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_{\sumtype}}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_{\prodtype}}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_{\prodtype}}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_{\prodtype}}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

Instead of spaces (aka discrete sets) consisting solely of
unstructured isolated points, we now investigate structured spaces
built from sums and products of elementary spaces. This structure
corresponds to the building blocks of type theory which are: the empty
type ($\bot$), the unit type ($\top$), the sum type ($\uplus$), and
the product ($\times$) type. Before getting into the formal theory,
let's consider possible deformations on the space $(\top \uplus \bot)
\times (\top \uplus \top)$. This space is the product of two
subspaces: the subspace $(\top \uplus \bot)$ which itself is the sum
of the space~$\top$ containing one element $\texttt{tt}$ and the empty
space $\bot$ and the subspace $(\top \uplus \top)$ which is the sum of
two spaces each containing the one element $\texttt{tt}$. First, as
discussed in the previous section, any deformation of this space must
at least preserve the number of elements: we can neither create nor
destroy points during any continuous deformation. Seeing that the
number of elements in our example space is 2, a reasonable hypothesis
is that we can deform the space above to any other space with 2
elements such as $\top \uplus \top$ or $\top \uplus (\top \uplus
\bot)$. What this really means is that we are treating the sum and
product structure as malleable. For example, imagining a product
structure as arranged in a grid; by ``stretching'' we can turn it
in to a sum structure arranged in a line. We can also change the
orientation of the grid by exchanging the axes, as well as do other
transformations --- as long as we preserve the number of points.
Of course, it is not a priori clear that this necessary requirement
is also sufficient.  Making this intuition precise will be the topic of this
section.

@@

\begin{center}
\begin{tabular}{c|c}
Logic & Types \\ \hline
$\false$ & $\bot$ \\
$\true$ & $\top$ \\
$\land$ & $\times$ \\
$\lor$ & $\uplus$ \\
\end{tabular}
\end{center}

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
type $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}, i.e. reduce to equivalent type-denoting
expressions under the rules of the host type system.  This is however
not true: the proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except in univalent type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into
$B \times A$, and vice-versa, which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

@@

@@

We are now in position to describe the commutative
semiring structure for types. After replacing the set $R$ with a
universe $U$, we also replace the algebraic use of $=$ in
Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
this change, we can indeed prove that types (with $\bot, \top, \uplus,
\times$) form a commutative semiring. The reader familiar with
universal algebra should pause and ponder a bit about what we have
done. We have lifted \emph{equality} from being in the signature of
the ambient logic and instead put it in the signature of the algebraic
structure of interest.  In simpler terms, we shift equality from
having a privileged status in our meta-theory, to being just another
symbol (denoting an equivalence relation) in our theory.  The understanding
that equality is not an absolute concept has recently been an area of
active research in mechanized mathematics --- although the concepts of
intensional versus extensional equality go back to Frege and Russell.

If we revisit the Curry-Howard correspondence, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow \qquad A \
  \text{is inhabited} \] The above is the real \textit{essence} of the
Curry-Howard correspondence.  In other words, classical Curry-Howard
tells us about \emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
however, they are not inverses.

%% \roshan{Use something else other than $*$ here to talk about
%%   idempotence.}

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$ versus
$\top$; the first has exactly \emph{two} proofs of inhabitation
($\inleft{\texttt{tt}}$ and $\inright{\texttt{tt}}$) while the second
only one ($\texttt{tt}$). These cannot be put in bijective
correspondence. Even though the ``payload'' \texttt{tt} is the same,
forgetting $\texttt{left}$ (or \texttt{right}) throws away information
-- something we have expressly disallowed.  Yes, this should remind
you of Maxwell's daemon: even though the data is the same, they are
tagged differently, and these tags are indeed information, and their
information content must be preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& A & \simeq & A &: \idc \\
\\
\identlp :&  \bot \uplus A & \simeq & A &: \identrp \\
\swapp :&  A \uplus B & \simeq & B \uplus A &: \swapp \\
\assoclp :&  A \uplus (B \uplus C) & \simeq & (A \uplus B) \uplus C &: \assocrp \\
\\
\identlt :&  \top \times A & \simeq & A &: \identrt \\
\swapt :&  A \times B & \simeq & B \times A &: \swapt \\
\assoclt :&  A \times (B \times C) & \simeq & (A \times B) \times C &: \assocrt \\
\\
\absorbr :&~ \bot \times A & \simeq & \bot ~ &: \factorzl \\
\dist :&~ (A \uplus B) \times C & \simeq & (A \times C) \uplus (B \times C)~ &: \factor
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

Nevertheless, the Curry-Howard correspondence still has some force. We
know that the inhabitants of types formed with with
$\bot, \top, \uplus, \times$ form a commutative semiring. What we want
to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal.  We have a
set of generators for those equations, namely those in
Def.~\ref{def:rig}. What we thus need is to create $8$ pairs of
mutually inverse functions which witness these identities.  For
concreteness, we show the signatures in Fig.~\ref{type-isos}.  As
these all come in symmetric pairs (some of which are self-symmetric),
we give names for both directions.

@@

We now have in our hands our desired denotational semantics for types.
We want to create a programming language, which we call $\Pi$, such
that the types and type combinators map to
$\bot, \top, \uplus, \times$, and such that we have ground terms whose
denotation are all $16$ type isomorphisms of
Fig.~\ref{type-isos}. This is rather straightforward, as we can simply
do this literally. To make the analogy with commutative semirings
stand out even more, we will use $0, 1, \sumtype$, and ${\prodtype}$ at the type level,
and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
shows the ``constants'' of the language. Note how we have continued
with the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof
  terms}, but rather than being witnesses of inhabitation, they are
witnesses of equivalences. Thus we get an unexpected programming
language design:

@@

Thus the denotation of the $\Pi$ terms \emph{should} be
permutations. But given types $A$ and $B$ denoting $\fin{m}$
and~$\fin{n}$ respectively, what are $A \uplus B$ and $A \times B$ ?
They correspond exactly to $\fin{m+n}$ and $\fin{m*n}$!
Geometrically, this corresponds to concatenation for $A + B$,
i.e. lining up the elements of $A$ first, and then those of~$B$. For
$A * B$, one can picture this as lining up the elements of $A$
horizontally, those of $B$ vertically and perpendicular to those of
$A$, and filling in the square with pairs of elements from $A$ and
$B$; if one re-numbers these sequentially, reading row-wise, this
gives an enumeration of $\fin{m*n}$.

@@

\begin{definition}($\sharp$). By recursion on $|b|$.  First define a
type associated to each size. We will revert to using type notation
for greater clarity of this definition:
\[\begin{array}{rcl}
  \tau~ (0) & = & \bot \\
  \tau~ (1 + n) & = & \top \uplus \tau~ (n) \\
 \end{array}\]
\noindent so that we can define $\sharp\, b = \tau~ |b|$.
\end{definition}

@@

\item
The product type \ensuremath{b_1 {\prodtype} b_2} may be represented both as one wire
labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

>>>>

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}

\newcommand{\presumtype}{\uplus}
\newcommand{\preprodtype}{*}
\newcommand{\sumtype}{+}
\newcommand{\prodtype}{\times}

\newcommand{\inleft}[1]{\textsf{left}~#1}
\newcommand{\inright}[1]{\textsf{right}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_{\sumtype}}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_{\sumtype}}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_{\sumtype}}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_{\prodtype}}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_{\prodtype}}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_{\prodtype}}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

Instead of spaces (aka discrete sets) consisting solely of
unstructured isolated points, we now investigate structured spaces
built from sums and products of elementary spaces. This structure
corresponds to the building blocks of type theory which are: the empty
type ($\bot$), the unit type ($\top$), the sum type ($\presumtype$), and
the product ($\preprodtype$) type. Before getting into the formal theory,
let's consider possible deformations on the space $(\top \presumtype \bot)
\preprodtype (\top \presumtype \top)$. This space is the product of two
subspaces: the subspace $(\top \presumtype \bot)$ which itself is the sum
of the space~$\top$ containing one element $\texttt{tt}$ and the empty
space $\bot$ and the subspace $(\top \presumtype \top)$ which is the sum of
two spaces each containing the one element $\texttt{tt}$. First, as
discussed in the previous section, any deformation of this space must
at least preserve the number of elements: we can neither create nor
destroy points during any continuous deformation. Seeing that the
number of elements in our example space is 2, a reasonable hypothesis
is that we can deform the space above to any other space with 2
elements such as $\top \presumtype \top$ or $\top \presumtype (\top \presumtype
\bot)$. What this really means is that we are treating the sum and
product structure as malleable. For example, imagining a product
structure as arranged in a grid; by ``stretching'' we can turn it
in to a sum structure arranged in a line. We can also change the
orientation of the grid by exchanging the axes, as well as do other
transformations --- as long as we preserve the number of points.
Of course, it is not a priori clear that this necessary requirement
is also sufficient.  Making this intuition precise will be the topic of this
section.

@@

\begin{center}
\begin{tabular}{c|c}
Logic & Types \\ \hline
$\false$ & $\bot$ \\
$\true$ & $\top$ \\
$\land$ & $\preprodtype$ \\
$\lor$ & $\presumtype$ \\
\end{tabular}
\end{center}

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
type $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}, i.e. reduce to equivalent type-denoting
expressions under the rules of the host type system.  This is however
not true: the proposition $A \preprodtype B \equiv B \preprodtype A$ is not
normally\footnote{Except in univalent type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \preprodtype B$ and $B \preprodtype A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \preprodtype B$ can be reversibly deformed into
$B \preprodtype A$, and vice-versa, which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

@@

%% \roshan{This is a poor sentence. We have 4 or 5 definitions back to
%%   back and have given very little intuition as to how or why they
%%   are relevant and then we go on to say this:}

@@

%% \roshan{More on the above: Let us either explain what the point is
%%   OR simply not suggest that we could be making a point, which we
%%   then choose not to.}

We are now in position to describe the commutative
semiring structure for types. After replacing the set $R$ with a
universe $U$, we also replace the algebraic use of $=$ in
Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
this change, we can indeed prove that types (with $\bot, \top, \presumtype,
\preprodtype$) form a commutative semiring. The reader familiar with
universal algebra should pause and ponder a bit about what we have
done. We have lifted \emph{equality} from being in the signature of
the ambient logic and instead put it in the signature of the algebraic
structure of interest.  In simpler terms, we shift equality from
having a privileged status in our meta-theory, to being just another
symbol (denoting an equivalence relation) in our theory.  The understanding
that equality is not an absolute concept has recently been an area of
active research in mechanized mathematics --- although the concepts of
intensional versus extensional equality go back to Frege and Russell.

If we revisit the Curry-Howard correspondence, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \presumtype A$ nor $A \preprodtype A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \presumtype A \ \text{is inhabited} \qquad \Leftrightarrow \qquad A \
  \text{is inhabited} \] The above is the real \textit{essence} of the
Curry-Howard correspondence.  In other words, classical Curry-Howard
tells us about \emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \presumtype A \rightarrow A$ and $g : A \rightarrow A \presumtype A$;
however, they are not inverses.

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $\circ$
where $\forall a. a \circ a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \presumtype \top$ versus
$\top$; the first has exactly \emph{two} proofs of inhabitation
($\inleft{\texttt{tt}}$ and $\inright{\texttt{tt}}$) while the second
only one ($\texttt{tt}$). These cannot be put in bijective
correspondence. Even though the ``payload'' \texttt{tt} is the same,
forgetting $\texttt{left}$ (or \texttt{right}) throws away information
-- something we have expressly disallowed.  Yes, this should remind
you of Maxwell's daemon: even though the data is the same, they are
tagged differently, and these tags are indeed information, and their
information content must be preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
& A & \simeq & A &\\
\\
&  \bot \presumtype A & \simeq & A &\\
&  A \presumtype B & \simeq & B \presumtype A &\\
&  A \presumtype (B \presumtype C) & \simeq & (A \presumtype B) \presumtype C &\\
\\
&  \top \preprodtype A & \simeq & A &\\
&  A \preprodtype B & \simeq & B \preprodtype A &\\
&  A \preprodtype (B \preprodtype C) & \simeq & (A \preprodtype B) \preprodtype C &\\
\\
& \bot \preprodtype A & \simeq & \bot &\\
& (A \presumtype B) \preprodtype C & \simeq & (A \preprodtype C) \presumtype (B \preprodtype C) &
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

%% \begin{figure}[t]
%% \[
%% \begin{array}{rrcll}
%% \idc :& A & \simeq & A &: \idc \\
%% \\
%% \identlp :&  \bot \presumtype A & \simeq & A &: \identrp \\
%% \swapp :&  A \presumtype B & \simeq & B \presumtype A &: \swapp \\
%% \assoclp :&  A \presumtype (B \presumtype C) & \simeq & (A \presumtype B) \presumtype C &: \assocrp \\
%% \\
%% \identlt :&  \top \preprodtype A & \simeq & A &: \identrt \\
%% \swapt :&  A \preprodtype B & \simeq & B \preprodtype A &: \swapt \\
%% \assoclt :&  A \preprodtype (B \preprodtype C) & \simeq & (A \preprodtype B) \preprodtype C &: \assocrt \\
%% \\
%% \absorbr :&~ \bot \preprodtype A & \simeq & \bot ~ &: \factorzl \\
%% \dist :&~ (A \presumtype B) \preprodtype C & \simeq & (A \preprodtype C) \presumtype (B \preprodtype C)~ &: \factor
%% \end{array}
%% \]
%% \caption{Type isomorphisms.}
%% \label{type-isos}
%% \end{figure}

Nevertheless, the Curry-Howard correspondence still has some force. We
know that the inhabitants of types formed with with $\bot, \top,
\presumtype, \preprodtype$ form a commutative semiring. What we want
to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal.  We have a
set of generators for those equations, namely those in
Def.~\ref{def:rig}. What we thus need is to create $8$ pairs of
mutually inverse functions which witness these identities.  For
concreteness, we show the signatures in Fig.~\ref{type-isos}.

@@

We now have in our hands our desired denotational semantics for types.
We want to create a programming language, which we call $\Pi$, such
that the types and type combinators map to $\bot, \top, \presumtype,
\preprodtype$, and such that we have ground terms whose denotation are
all $16$ type isomorphisms of Fig.~\ref{type-isos}. This is rather
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, \sumtype$, and ${\prodtype}$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language.  As these all come in symmetric pairs
(some of which are self-symmetric), we give names for both directions.
Note how we have continued with the spirit of Curry-Howard: the terms
of $\Pi$ are \emph{proof terms}, but rather than being witnesses of
inhabitation, they are witnesses of equivalences. Thus we get an
unexpected programming language design:

@@

Thus the denotation of the $\Pi$ terms \emph{should} be
permutations. But given types $A$ and $B$ denoting $\fin{m}$
and~$\fin{n}$ respectively, what are $A \presumtype B$ and $A \preprodtype B$ ?
They correspond exactly to $\fin{m+n}$ and $\fin{m*n}$!
Geometrically, this corresponds to concatenation for $A + B$,
i.e. lining up the elements of $A$ first, and then those of~$B$. For
$A * B$, one can picture this as lining up the elements of $A$
horizontally, those of $B$ vertically and perpendicular to those of
$A$, and filling in the square with pairs of elements from $A$ and
$B$; if one re-numbers these sequentially, reading row-wise, this
gives an enumeration of $\fin{m*n}$.

@@

\begin{definition}($\sharp$).
  By recursion on $|b|$.  First define $\tau$ that maps numeric sizes
  to their corresponding types. We will revert to using type notation
  for greater clarity of this definition:
\[\begin{array}{rcl}
  \tau~ (0) & = & \bot \\
  \tau~ (1 + n) & = & \top \presumtype \tau~ (n) \\
 \end{array}\]
\noindent so that we can define $\sharp\, b = \tau~ |b|$.
\end{definition}

@@

\item
The product type \ensuremath{b_1 {\prodtype} b_2} may be represented both as one wire
labeled \ensuremath{b_1 \prodtype b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.
